---
title: "Using metabalize"
author: "Andrew D. Steen"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using metabalize}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

```{r}
require(ggplot2)
require(plyr)
require(reshape2)

raw_list <- read_MAVEN(data_fn="../data/140114_AET_13C_glucose_flux.csv",
                  key_fn="../data/curacao_samplekey.csv")

# Split the data from the experimental variables metadata
#    I should probably change the exp.var workflow
raw <- raw_list$raw_data
exp.var <- raw_list$exp.var


# Convert replicate variable into factor - not sure how
raw$replicate <- as.factor(raw$replicate)

################
# Create data frame of only 12C isotopomers; expressed as fraction of total ions 
################

# calc_12C identifies 12C ions
system.time({
  raw_12C <- calc_12C(raw) 
}) # like 5 seconds on my macbook air

# Filter 12C peaks from 13C-containing peaks
C12_only <- subset(raw_12C, is.12C==TRUE)

################
# Fit exp models to each unique sample, metabolite, treatment & sample.type
################

# # Determine the number of timepoints for each sample, compound, treatment, replicate and sample type
# n_timepoints <- ddply(C12_only, c("compound", "treatment", "replicate", "sample.type"), 
#                      summarise,
#                      n_tp=length(time))
# 
# # Actually create a model for each sample, compound, treatement, replicate and sample type

min_for_debugging <- C12_only[ , c("compound", "treatment", "sample.type", "time", "relative.ion.count")]
system.time({ # This throws an error - I don't know why
  nls_fits <- dlply(min_for_debugging, #C12_only,
                    c("compound", "treatment", "sample.type"), safe_NLS, .inform=TRUE)
}) # ~4 seconds on my laptop, not so bad

# How many of these are good?
bad_fits <- ldply(nls_fits, is.null)
sum(bad_fits$V1)/nrow(bad_fits) #Now none seem to be bad
```


# Getting the data out
```{r}
# Get the nls coefficients out
all_coefs <- ldply(nls_fits, function(x) exp_coef(x))
print(all_coefs)

# How many of the A values are good?
ggplot(all_coefs, aes(x=A.est)) + 
  geom_density() #Many are close to zero

# Create predicted values for exponential fit
all_preds <- ldply(nls_fits, safe_NLS_pred, c(0, 125))

# Plot all the data
ggplot() +
  geom_point(data=C12_only, aes(x=time, y=relative.ion.count, colour=sample.type)) + 
  geom_line(data=all_preds, aes(x=time, y=relative.ion.count, colour=sample.type)) + 
  facet_wrap(~compound)
ggsave("../testplot.png", height=20, width=20, units="in", dpi=300, type="cairo")
p <- plot_timecourse(C12_only, 
                     exp.var=c("compound", "treatment", "sample.type", "time"),
                     exp_pred=all_preds)
print(p)

```

Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` setion of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
